---
title: Hunting for Attractors
author: ''
date: '2020-01-05'
slug: hunting-for-attractors
categories: []
tags:
  - R
  - generative-graphics
  - visualization
---

I found this wonderful [book by Julien C. Sprott](http://sprott.physics.wisc.edu/sa.htm) about [strange attractors](https://en.wikipedia.org/wiki/Attractor#Strange_attractor), free to download no less! Not only is the book full of beautiful images, is also has lots of source code to generate them. A small thing though: the book is from 1995^[1995 happens to be when my grandmother bought me my first computer: a Pentium 133 MHz with Windows 95 on it. I learned about fractals roughly at the same time, so reading this book is quite nostalgic.], and the code is written in BASIC. I decided to recreate it in R.

In the book's second chapter, I learned about a neat trick to sift through the huge parameter spaces of these attractors to find beautiful patterns. I can highly recommend the book if you are interested in more details or rigor, as I will only mention what is necessary to take us to pretty pixels.

# Strange attractors

Playing slightly loose with the definitions, an image of a strange attractor is made by generating a sequence of points from an iterated function ("map"), which are then plot. In this post, I will generate pictures from a "two dimensional quadratic map". This means that I will run through the functions below with different parameters for a and starting points x and y:

$$
x_{n+1} = a_1 + a_2x_n + a_3x_n^2 + a_4x_ny_n + a_5y_n + a_6y_n^2\\
y_{n+1} = a_7 + a_8x_n + a_9x_n^2 + a_{10}x_ny_n + a_{11}y_n + a_{12}y_n^2
$$

To see a first pretty picture, we choose some suitable values for the 12 parameters (more about this in a bit), generate bunch of steps and plot them.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
theme_set(theme_void() + theme(legend.position = 'none'))

# I will go parallel a little later; set up my computer for this
library(future)
library(furrr)
plan(multiprocess)
```

First, `quadratic_2d_step()` takes the 12 parameters and an x/y pair to generate the next step.

```{r}
quadratic_2d <- function(a, xn, yn) {
  xn1 <- a[1] + a[2]*xn + a[3]*xn*xn + a[ 4]*xn*yn + a[ 5]*yn + a[ 6]*yn*yn
  yn1 <- a[7] + a[8]*xn + a[9]*xn*xn + a[10]*xn*yn + a[11]*yn + a[12]*yn*yn
  
  c(xn1, yn1)
}
```

We also need a function to run through these iterations and collect the results. In return from `iterate()`, we get a list of visited points by the iteration. As an example, I use the `a` parameters to generate the [Hénon map](https://en.wikipedia.org/wiki/H%C3%A9non_map).

```{r}
iterate <- function(step_fn, a, x0, y0, iterations) {
  x <- rep(x0, iterations)
  y <- rep(y0, iterations)
  
  for(n in 1:(iterations - 1)) {
    xy <- step_fn(a, x[n], y[n])
    x[n+1] <- xy[1]
    y[n+1] <- xy[2]
  }
  
  tibble(x = x, y = y) %>%
    mutate(n = row_number())
}

henon_a <- c(1, 0, -1.4, 0, 0.3, 0, 0, 1, 0, 0, 0, 0)
iterate(step_fn = quadratic_2d, a = henon_a, x0 = 0, y0 = 0, iterations = 10)
```

Let's generate a larger number of points, and see the result.


```{r}
iterate(step_fn = quadratic_2d, a = henon_a, x0 = 0, y0 = 0, iterations = 5000) %>%
  ggplot(aes(x, y)) +
    geom_point(size = 0, shape = 20, alpha = 0.2) + # Plot a pixel for each point
    coord_equal()
```

There we go! We have a first image.

# Looking for chaos

Remember that a 2D quadratic map has 12 (!) parameters to pick. Not all sets give interesting results.

There are three broad categories of parameter sets:

 * Converging: these are plots where the point end up in a fixed point, regardless of where they start. Nothing much to see here.
 * Run-away: Here, the x/y pair runs away to infinity. Again, nothing much to see here.
 * Chaotic: These are the ones we're looking for!

Let's assume that any value between, say, -1.5 and 1.5 with 0.01 step increments generates a unique image. This means that we have `301^12 = 10^70` different charts to sift through. Of these, roughly a single digit percentage are chaotic and interesting. We'd like the computer to do the tedious work of discarding those that are boring.

This, we can tell a computer to do. We can think of a chaotic system as having these characteristics:

 * Points try to run away from each other, as opposed to settle down. This removes the converging scenario.
 * Despite trying to run away, the points stay within finite bounds. This removes the run-away scenario.
 
An algorithm for this is: Follow a point `x/y`. Before taking a step, place a second point just next to this one. Let both points step one iteration forward, and check whether the distance between the two points have increased or decreased. Do this a fair number of times, and check whether the average ratio of the distance `after / before` is more than 1. If so, the points try to increase the distance between them.

To mathematize this a little more: calculate

$$L = \sum log_2 \dfrac{d_{n+1}}{d_n} / N$$

where d is the distance between the two points and N is the number of points evaluated. Since we're using log:ed values^[Averaging log values means that we're calculating the geometrical mean], the threshold value where looking for is L = 0 (`2^0 = 1`).

Now, if L > 0 and the x/y values are not growing without bound (say, within +/- a million), we have a set of parameters that just might be chaotic.

First up, let's implement a function to calculate L.

```{r}
L <- function(step_fn, a, x0, y0, iterations = 1000) {
  # Really, put the point nearby and see what happens
  nearby_distance <- 0.000001
  
  xy <- c(x0, y0)
  xy_near <- xy + c(nearby_distance, 0)
  
  # Collect the log distance ratios as we iterate
  sum_log_distance_ratios <- 0
  
  for (n in 1:iterations) {
    xy <- step_fn(a, xy[1], xy[2])
    xy_near <- step_fn(a, xy_near[1], xy_near[2])
    
    new_distance = sqrt((xy[1] - xy_near[1])^2 + (xy[2] - xy_near[2])^2)
    
    if (new_distance == 0) {
      # The points have converged
      return (-Inf)
    }
    
    if (abs(new_distance) == Inf) {
      # The points have run away
      return (Inf)
    }
    
    # Move near point after xy
    # We put the near point just behind in the direction that they differ
    angle = atan2(xy_near[2] - xy[2], xy_near[1] - xy[1])
    xy_near <- c(xy[1] + nearby_distance * cos(angle),
                 xy[2] + nearby_distance * sin(angle))
    
    
    sum_log_distance_ratios = sum_log_distance_ratios + log2(new_distance / nearby_distance)
    
  }
  
  sum_log_distance_ratios / iterations
}
```

Let's try it on a few sets of parameters. The Hénon map is chaotic, remember?

```{r}
L(quadratic_2d, henon_a, 0.01, 0.01)
```

With all zeros, the process settles at (0, 0).

```{r}
L(quadratic_2d, rep(0, 12), 0.01, 0.01)
```



# Generating candidates

All moving parts are now in place. Let's randomize a bunch of parameters, filter out those we believe are chaotic, and see for ourselves:

```{r}
set.seed(1)

df <- tibble(pattern = 1:1000) %>%
  mutate(a = map(pattern, ~ round(runif(12, -1.5, 1.5), 2))) %>%
  mutate(L_val = map_dbl(a, ~ L(quadratic_2d, ., 0, 0))) %>%
  filter(L_val > 0)

print(df)
```

Here, we get `r nrow(df)` candidate parameter sets with an L value > 0. Some of these might run away later, but we only need to generate and plot full sequences for these few. We'll also `normalize()` the x/y points -- put them in a range from 0 to 1 -- to make them plot on the same scales.

```{r}
normalize_xy <- function(df) {
  range <- with(df, max(max(x) - min(x), max(y) - min(y)))
  
  df %>%
    mutate(x = (x - min(x)) / range,
           y = (y - min(y)) / range)
  
}

render_grid <- function(df, iterations) {
  df %>%
    mutate(xy = map(a, ~ iterate(quadratic_2d, ., 0, 0, iterations))) %>%
    
    # Remove those who have grown very large / might run away
    filter(map_lgl(xy, function(d) with(d, all(abs(x) + abs(y) < 1e7)))) %>%
    
    mutate(xy = map(xy, normalize_xy)) %>%
    
    unnest(xy) %>%
    ggplot(aes(x, y)) +
      geom_point(size = 0, shape = 20, alpha = 0.1) +
      coord_equal() +
      facet_wrap(~ pattern)
}

df %>%
  render_grid(5000)
```



There we go! A few of these have quite boring periodic patterns, not really fit for framing. Let's render one in higher quality anyways:


```{r}
df %>%
  filter(pattern == 423) %>%
  render_grid(100000)
```

Let's not spend too much time on these ones. There are a few that look promising in this haul:


```{r}
df %>%
  filter(pattern == 683) %>%
  render_grid(100000)
```

```{r}
df %>%
  filter(pattern == 694) %>%
  render_grid(100000)
```


```{r}
df %>%
  filter(pattern == 75) %>%
  render_grid(100000)
```

Amazing pictures from such simple math!

# Generating print quality

Finally, we can tweak the rendering of whatever patterns we find for even more effects. One trick to use is to not plot each point individually, but to "bin" them on a grid, counting how many points are in each cell. This number can then be mapped to color and transparency values. I'll build a separate function for this.

To speed things up slightly, I'll parallelize the data generation. I will divide the iterations equally among the cores in my CPU. I'll select slightly different starting points for each thread. This gives eight (or however many cores I happen to have) separate iterations, but this won't show since the attractors by definition attract similar paths.

```{r}
render_print_data <- function(a, iterations, gridsize) {
  CPU_cores <- parallel::detectCores()
  
  tibble(thread = 1:CPU_cores) %>%
    mutate(x0 = runif(length(thread), -0.1, 0.1),
           y0 = runif(length(thread), -0.1, 0.1)) %>%
    mutate(xy = future_pmap(., function(x0, y0, ...) iterate(quadratic_2d, a, x0, y0, iterations / CPU_cores))) %>%
    unnest(xy) %>%
    
    # Normalize
    mutate(range = max(max(x) - min(x), max(y) - min(y))) %>%
    mutate(x = (x - min(x)) / range,
           y = (y - min(y)) / range) %>%
    
    group_by(x = round(x * gridsize) / gridsize,
             y = round(y * gridsize) / gridsize) %>%
    summarize(n = n())
}

df.chosen <- df %>%
  filter(pattern == 694)

d <- render_print_data(df.chosen$a[[1]], 2000000, 1000)

print(head(d))
```


Finally, I can plot and style this data with all kinds of interesting visual choices. Try transforming the `n` variable for more or less dense pixels in the alpha value via for example `log`, `n^0.5`, or `n^0.3`. Trial-and-error seems to be the way to go here.

```{r}
d %>%
  ggplot(aes(x, y)) +
    geom_point(aes(alpha = sqrt(n), color = log(n)), size = 0, shape = 20) +
    scale_alpha_continuous(range = c(0, 1), limits = c(0, NA)) +
    scale_color_distiller(palette = 'YlOrRd', direction = 1) +
    coord_equal()
```



# Find something to frame

With this in hand, I can now set out to find many more patterns. I spent hours in the following process:

 1. Generate a grid of candidates with 5000 iterations, picking out those that seem promising
 2. Zoom in on the candidates using 100k iterations
 3. Take the ones I really like, render with a couple of million points and style until I'm happy.
 
Here's a few I liked!

```{r}
d <- render_print_data(a = c(-0.38, -0.60,  0.20, -0.79,  0.10,  0.54,  0.41, -0.59,  0.95, -1.19, -0.65, -0.66),
                  iterations = 4000000,
                  gridsize = 1000)

d %>%
  ggplot(aes(x, y)) +
    geom_point(aes(alpha = sqrt(n), color = log(n)), size = 0, shape = 20) +
    scale_alpha_continuous(range = c(0, 1), limits = c(0, NA)) +
    scale_color_gradientn(colors = lapply(colourlovers::clpalette(292482)$colors, function(c) paste0('#', c))) +
    coord_equal()

```



```{r}
d <- render_print_data(a = c(0.23, -0.15, -0.25, 0.71, 0.79, -0.04, -0.57, 0.91, 0.59, -0.9, -0.58, 1.02),
                  iterations = 4000000,
                  gridsize = 1000)

d %>%
  ggplot(aes(x, y)) +
    geom_point(aes(alpha = n ^ 0.05, color = log(n)), size = 0, shape = 20) +
    scale_alpha_continuous(range = c(0, 1), limits = c(0, NA)) +
    scale_color_gradientn(colors = lapply(colourlovers::clpalette(444487)$colors, function(c) paste0('#', c))) +
    coord_equal()

```





```{r}
d <- render_print_data(a = c(-0.16, -0.37, -0.27, 0.16, -0.66, -0.74, -1.11, -0.51, 0.59, 0.81, -0.06, -0.44),
                  iterations = 2000000,
                  gridsize = 1000)

d %>%
  ggplot(aes(x, y)) +
    geom_point(aes(alpha = sqrt(n), color = sqrt(n)), size = 0, shape = 20) +
    scale_alpha_continuous(range = c(0, 1), limits = c(0, NA)) +
    scale_color_gradientn(colors = rev(lapply(colourlovers::clpalette(131576)$colors, function(c) paste0('#', c)))) +
    coord_equal()

```


```{r}
d <- render_print_data(a = c(1.03, -1.11, 0.25, -0.51, 0.6, 0.33, 1.2, 0.28, -0.71, -0.68, -0.66, -0.8),
                  iterations = 2000000,
                  gridsize = 1000)

d %>%
  ggplot(aes(x, y)) +
    geom_point(aes(alpha = n^0.3, color = log(n)), size = 0, shape = 20) +
    scale_alpha_continuous(range = c(0, 1), limits = c(0, NA)) +
    scale_color_gradientn(colors = rev(lapply(colourlovers::clpalette(41095)$colors, function(c) paste0('#', c)))) +
    coord_equal()

```



```{r}
d <- render_print_data(a = c(0.94, -0.68, 0.54, -0.36, 0.42, -0.99, -0.92, -0.26, 0.82, -0.08, -0.48, 0.98),
                  iterations = 2000000,
                  gridsize = 1000)

d %>%
  ggplot(aes(x, y)) +
    geom_point(aes(alpha = n^0.65, color = sqrt(n)), size = 0, shape = 20) +
    scale_alpha_continuous(range = c(0, 1), limits = c(0, NA)) +
    coord_equal()

```











