---
title: Generative gravel
author: ''
date: '2020-01-02'
slug: generative-gravel
categories: []
tags:
  - generative-graphics
  - visualization
  - R
---

I [just learned](https://quantixed.org/2019/05/26/turn-a-square-generative-art/) about [Georg Nees](https://en.wikipedia.org/wiki/Georg_Nees), who was among the first people to create art using computers. I've never gotten deep into generative graphics, but find myself coming back to it every so often. Inspired by [Nees' 1968 piece Schotter](http://www.medienkunstnetz.de/works/schotter/) (en: Gravel / Crushed Stone) I wanted to try my hand at manipulating squares using R.

```{r message=FALSE, warning=FALSE}
library(tidyverse)

theme_set(theme_void() + theme(legend.position = 'none'))
```

Here's the plan: Create a 10 columns / 20 rows grid of squares. The top squares should remain steady and organized. As we go down the rows, we rotate and displace them more and more.


First out: the `square()`. I create a simple data frame with one row containing x/y coordinates for each of the four corners.

I'll later use `geom_polygon` to draw the shape:

  1) `geom_polygon` will close the shape, i.e. draw a final segment from the end back to the start. To use `geom_path` I would need the starting point a second time, as a fifth corner.
  2) `geom_polygon`s can be filled, with colors.
  

```{r}
square <- function() {
  tibble(x = 0.5 * c(1, -1, -1, 1),
         y = 0.5 * c(1, 1, -1, -1))
}

square() %>%
  ggplot(aes(x, y)) +
    geom_polygon(fill = NA, color = 'black') +
    coord_equal()

```


Next for a grid of squares. I want to rotate the points around the center of the squares. The center is at the origin `(0, 0)` when I create a square. The code is a fair bit easier if I rotate it around the origin. Therefore, my plan is to first create all the squares at `(0, 0)`, rotate them in place, and spread them over the grid only once the rotations are done.

```{r}
make_grid <- function(columns, rows) {
  tibble(square = 1:(columns * rows) - 1) %>% # Give each square a unique ID
                                              # Starting at 0 which simplifies the
                                              # calculations of column and row below
    mutate(points = map(square, ~ square())) %>%
    unnest(points) %>%
    mutate(column = square %% columns,
           row = square %/% columns)
}

spread_squares <- function(df, padding = 0.0) {
  df %>%
    mutate(x = x + column * (1 + padding),
           y = y + row * (1 + padding))
}

make_grid(10, 20) %>%
  spread_squares() %>%
  print() %>%
  ggplot(aes(x, y, group = square)) +
    geom_polygon(fill = NA, color = 'black') +
    coord_equal()
```


It looks like Nees did two transformations of the squares:

  1) Rotate the squares
  2) Move the squares

My rotation works by giving each corner an angle which it should rotate. I create a separate function to `rotate()` which takes a [tidy evaluation parameter](https://www.tidyverse.org/blog/2019/06/rlang-0-4-0/) for angles. This lets me use either values or formulas based on columns in the data to rotate.


```{r}
rotate <- function(df, angle) {
  df %>%
    mutate(xnew = x * cos({{angle}}) - y * sin({{angle}}),
           ynew = x * sin({{angle}}) + y * cos({{angle}}),
           x = xnew,
           y = ynew) %>%
    select(-xnew, -ynew)
}

# Rotate the squares 45 degrees = pi / 4 radians

make_grid(10, 20) %>%
  rotate(pi / 4) %>%
  spread_squares() %>%
  ggplot(aes(x, y, group = square)) +
    geom_polygon(fill = NA, color = 'black') +
    coord_equal()
  
```


## Schotter re-schottered

Now, all of the moving pieces are ready to recreate the picture.

I'll index the topmost row for zero and make sure it plots correctly by `scale_y_continuous(trans = 'reverse')`. `noisiness` is our variable to determine how much noise to add by square. In this piece, it is simply how far down the rows we've come.

Since I want the square to remain intact when I rotate and move around, I need to move all its four corners together. For this, I `group_by` the square id and draw a single random number for all its corners at once.

I use normally distributed random numbers with increasing standard deviations. I did mine slighly less noisy than the original, in particular so for the displacement.

```{r}
set.seed(1)

df.result <- make_grid(10, 20) %>%
  mutate(noisiness = row / max(row)) %>%
  
  group_by(square) %>%
  mutate(angle = rnorm(1, 0, noisiness * (pi / 2))) %>%
  ungroup() %>%
  
  rotate(angle) %>%
  spread_squares(padding = 0.2) %>%
  
  group_by(square) %>%
  mutate(x = x + rnorm(1, 0, noisiness ^ 2 * 0.3),
         y = y + rnorm(1, 0, noisiness ^ 2 * 0.3)) %>%
  ungroup()

df.result %>%
  ggplot(aes(x, y, group = square)) +
    geom_polygon(fill = NA, color = 'black') +
    coord_equal() +
    scale_y_continuous(trans = 'reverse')
```

Finally, I add [a little color](https://www.colourlovers.com/palette/292482/Terra) and call it done!

```{r}
df.result %>%
  ggplot(aes(x, y, group = square)) +
    geom_polygon(aes(fill = row), color = '#774F38', size = 0.5) +
    coord_equal() +
    scale_y_continuous(trans = 'reverse') +
    scale_fill_gradient(low = '#fcf0de', high = '#033649') +
    theme(panel.background = element_rect(fill = '#E8DDCB'),
          plot.background = element_rect(fill = '#E8DDCB'))
```


There's all kinds of cool things you can try with this: Try polygons other than squares, try rotations by something other than a straight gradient from little to much, or try to transform each corner individually to gradually distort the squares as they fall.

